---
title: "Lab 11 - Interactive Visualization"
output: html_document
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(eval = TRUE, include  = TRUE)
```

# Learning Goals

- Read in and process Starbucks data.
- Create interactive visualizations of different types using `plot_ly()` and `ggplotly()`.
- Customize the hoverinfo and other plot features.
- Create a Choropleth map using `plot_geo()`.

# Lab Description

We will work with two Starbucks datasets, one on the store locations (global) and one for the nutritional data for their food and drink items. We will do some text analysis of the menu items.

# Deliverables

Upload an html file to Quercus and make sure the figures remain interactive.

# Steps

### 0. Install and load libraries

```{r message=FALSE, echo=FALSE, warning=FALSE}

library(tidyverse)
library(plotly)
library(widgetframe)
library(tidytext)

```

### 1. Read in the data

- There are 4 datasets to read in, Starbucks locations, Starbucks nutrition, US population by state, and US state abbreviations. All of them are on the course GitHub.

```{r}

sb_locs <- read_csv("https://raw.githubusercontent.com/JSC370/JSC370-2025/refs/heads/main/data/starbucks/starbucks-locations.csv")

sb_nutr <- read_csv("https://raw.githubusercontent.com/JSC370/JSC370-2025/refs/heads/main/data/starbucks/starbucks-menu-nutrition.csv")

usa_pop <- read_csv("https://raw.githubusercontent.com/JSC370/JSC370-2025/refs/heads/main/data/starbucks/us_state_pop.csv")

usa_states<-read_csv("https://raw.githubusercontent.com/JSC370/JSC370-2025/refs/heads/main/data/starbucks/states.csv")
```


### 2. Look at the data

- Inspect each dataset to look at variable names and ensure it was imported correctly.


```{r}
head(sb_locs)
head(sb_nutr)
head(usa_pop)
head(usa_states)
```


### 3. Format and merge the data

- Subset Starbucks data to the US.
- Create counts of Starbucks stores by state.
- Merge population in with the store count by state.
- Inspect the range values for each variable. 

```{r}

sb_usa <- sb_locs |> filter(Country == "US")

sb_locs_state <- sb_usa |>
  rename(state = 'State/Province') |>
  group_by(state) |>
  summarize(n_stores = n())

# need state abbreviations
usa_pop_abbr <- 
  full_join(usa_pop, usa_states,
            by = join_by(state == State) 
            ) 
  
sb_locs_state <- full_join(sb_locs_state, usa_pop_abbr,
                           by = join_by(state == Abbreviation)
                           )
  


```


### 4. Use `ggplotly` for EDA

Answer the following questions:

- Are the number of Starbucks proportional to the population of a state? (scatterplot)

- Is the caloric distribution of Starbucks menu items different for drinks and food? (histogram)

- What are the top 20 words in Starbucks menu items? (bar plot)

- 4a) Answer: 
```{r}
p1 <- ggplot(sb_locs_state, aes(x = population, y = n_stores, colour = state)) +
  geom_point(alpha = 0.8) +
  theme_bw()

ggplotly(p1)
```

The number of Starbucks stores generally increases with a state's population, but it’s not perfectly proportional. Some states, like California, have way more stores than expected, likely due to factors like urbanization and demand.

- 4b) Answer: 

```{r}
p2 <- ggplot(sb_nutr, aes(x=Calories, fill=Category)) + 
  geom_histogram(alpha= 0.5) + 
  theme_bw()

ggplotly(p2)
```

The caloric distribution of Starbucks menu items is different for drinks and food. Drinks tend to have lower calorie counts and peak around 150 calories, while food items generally have higher calorie counts and peak around 450 calories.

- 4c) Answer: 

```{r}
p3 <- sb_nutr |>
  unnest_tokens(word, Item, token = "words") |>
  count(word, sort = T) |>
  head(20) |>
  ggplot(aes(fct_reorder(word, n), n)) +
  geom_col() +
  coord_flip() +
  theme_bw()

ggplotly(p3)
```

The top 20 most common words in Starbucks menu items include "iced," "tazo," "bottled," "sandwich," "chocolate, "coffee", and "tea", with "iced" appearing the most frequently. This suggests a strong emphasis on cold drinks, coffees, cholocate, and popular food items like sandwiches and desserts.


### 5. Scatterplots using `plot_ly()`

- Create a scatterplot using `plot_ly()` representing the relationship between calories and carbs. Color the points by category (food or beverage). Is there a relationship, and do food or beverages tend to have more calories?

```{r}
sb_nutr |>
  plot_ly(x = ~Calories, y = ~`Carb. (g)`,
          type = 'scatter', mode = 'markers', color = ~Category)
```

- 5a) Answer:
There is a positive relationship between calories and carbs—items with more calories tend to have more carbs. Food items generally have higher calories than beverages, while drinks cluster more at lower calorie values.

- Repeat this scatterplot but for the items that include the top 10 words. Color again by category, and add hoverinfo specifying the word in the item name. Add layout information to title the chart and the axes, and enable `hovermode = "compare"`.
- What are the top 10 words and is the plot much different than above?
  
```{r}
topwords <- sb_nutr |>
  unnest_tokens(word, Item, token = "words") |>
  group_by(word) |>
  summarise(word_frequency = n()) |>
  arrange(across(word_frequency, desc)) |>
  head(10)

sb_nutr |>
  unnest_tokens(word, Item, token="words") |>
  filter(word %in% topwords$word) |>
  plot_ly(x = ~Calories, y = ~`Carb. (g)`,
          type = 'scatter', mode = 'markers',
          color = ~Category,
          hoverinfo = 'text',
          text = ~paste0("Item: ", word)
  ) |>
  layout(
    title = 'Cal vs Carbs', 
    xaxis = list(title = 'Calories'),
    yaxis = list(title = 'Carbs'),
    hovermode = 'compare'
  )
```

- 5b) Answer:
The filtered scatterplot should look similar but will have fewer points, only showing items with the top 10 words. Since many common menu items contain these words (especially drinks like "Iced Coffee" or "Tazo Tea"), the trend between calories and carbs will still be visible, but it may have a stronger emphasis on popular drink and food categories.


### 6. `plot_ly` Boxplots

- Create a boxplot of all of the nutritional variables in groups by the 10 item words.
- Which top word has the most calories? Which top word has the most protein?

```{r}
sb_nutr_long <- sb_nutr |>
  unnest_tokens(word, Item, token="words") |>
  filter(word %in% topwords$word) |>
  pivot_longer(
    cols = c(Calories, `Fat (g)`, `Carb. (g)`, `Fiber (g)`, `Protein (g)`),
    names_to = "Nutrient", values_to = "Value")

plot_ly(data = sb_nutr_long,
        x = ~word,
        y = ~Value,
        color = ~Nutrient,
        type = 'box'
  ) |>
  layout(
    title = "Nutrition values for the top 10 words items",
    xaxis = list(title = 'Item Word'),
    yaxis = list(title = 'Nutrition Value'),
    hovermode = 'compare'
  )
```

- 6) Answer:
The top word with the most calories is "sandwich", however, it has a high variability since the IQR is quite large. On the other hand, the top word with the most protein is egg, which makes sense.

### 7. 3D Scatterplot

- Create a 3D scatterplot between Calories, Carbs, and Protein for the items containing the top 10 words
- Do you see any patterns (clusters or trends)?

```{r}
sb_nutr |>
  unnest_tokens(word, Item, token = "words") |>
  filter(word %in% topwords$word) |>
  plot_ly(
    x = ~Calories,
    y = ~`Carb. (g)`,
    z = ~`Protein (g)`,
    color = ~word,
    type = 'scatter3d',
    mode = 'markers',
    marker = list(size = 5)
  ) |>
  layout(
    title = "3D Scatterplot of Calories, Carbs, and Protein",
    scene = list(
      xaxis = list(title = "Calories"),
      yaxis = list(title = "Carbohydrates (g)"),
      zaxis = list(title = "Proein (g)")
    )
  )
```

- 7) Answer:
The 3D scatterplot shows that items with higher calories generally have higher carbs and protein, but distinct clusters exist. Egg and sandwich items have the most protein, while chocolate, bottled, and tea items are higher in carbs but lower in protein. Drinks like black coffee and iced tea tend to be low in all three nutrients.

### 8. `plot_ly` Map

- Create a map to visualize the number of stores per state, and another for the population by state. Add custom hover text. Use subplot to put the maps side by side.
- Describe the differences if any.

```{r}
# Set up mapping details
set_map_details <- list(
  scope = 'usa',
  projection = list(type = 'albers usa'),
  showlakes = TRUE,
  lakecolor = toRGB('steelblue')
)

# Make sure both maps are on the same color scale
shadeLimit <- 125

# Create hover text
sb_locs_state$hover <- with(sb_locs_state, paste("Number of Starbucks: ", n_stores, '<br>', "State: ", state.y, '<br>', "Population: ", population))

# Create the map
map1 <- plot_geo(sb_locs_state, locationmode = "USA-states") |>
  add_trace(z = ~n_stores, text = ~hover, locations = ~state,
            color = ~n_stores, colors = 'Purples') |>
  layout(title = "Starbucks store by state", geo = set_map_details)
map1


map2 <- plot_geo(sb_locs_state, locationmode = "USA-states") |>
  add_trace(z = ~population, text = ~hover, locations = ~state,
            color = ~population, colors = 'Purples') |>
  layout(title = "Population by state", geo = set_map_details)
map2

subplot(map1, map2) |>
    layout(
    annotations = list(
      list(x = 0.2, y = 1.0, text = "Starbucks Stores by State", showarrow = FALSE, xref = 'paper', yref = 'paper', font = list(size = 14, color = "black")),
      list(x = 0.8, y = 1.0, text = "Population by State", showarrow = FALSE, xref = 'paper', yref = 'paper', font = list(size = 14, color = "black"))
    )
  )
```

- 8) Answer:
The left map shows the number of Starbucks stores per state, while the right map shows the population of each state. While California has the most stores and the largest population, some states (like Texas and Florida) have high populations but fewer stores than expected. This suggests that while population size influences store count, other factors like urbanization, demand, and market strategy also play a role.